---
title: "Statistical Rethinking -- Chapters 3"
output: html_document
---


```{r setup, include=FALSE}
library(rethinking)
library(kableExtra)
library(tibble)
library(dplyr)
```

## p49
```{r}
p_vampire = 0.001
p_positive_vampire = 0.95
p_positive_mortal = 0.01

p_positive = p_vampire * p_positive_vampire + (1 - p_vampire) * p_positive_mortal

p_vampire_positive = p_positive_vampire * p_vampire / p_positive

p_vampire_positive
```

## p52-54
```{r}
num_points = 1000
N = 9
x = 6

globe_toss_uniform = function(x, N, num_points) {
  p_grid = seq(from=0, to=1, length.out=num_points)
  prior = rep(1, num_points)
  likelihood = dbinom(x, size=N, prob=p_grid) 
  posterior = likelihood * prior
  posterior = posterior / sum(posterior)
  return(list(p_grid=p_grid, posterior=posterior))
}

result = globe_toss_uniform(x, N, num_points)
p_grid = result$p_grid
posterior = result$posterior
plot(x=p_grid, y=posterior)
```

```{r}
num_samples = 1e4
samples = sample(p_grid, prob=posterior, size=num_samples, replace=T)
plot(samples)
```

```{r}
dens(samples)
```

### Intervals of defined boundaries
```{r}
sum(posterior[p_grid < 0.5])
```

```{r}
mean(samples < 0.5)
```

```{r}
mean(samples > 0.5 & samples < 0.75)
```

## p.55
### Intervals of defined mass

```{r}
quantile(samples, 0.8)
```

```{r}
quantile(samples, c(0.1, 0.9))
```

```{r}
result = globe_toss_uniform(3, 3, num_points)
plot(x=result$p_grid, y=result$posterior)
```

```{r}
samples = sample(x=result$p_grid, prob=result$posterior, size=num_samples, replace=T)
quantile(samples, c(0.25, 0.75))
```

```{r}
# TODO: find how this method is implemented
HPDI(samples, 0.5)
```

## p.59
### Point estimates
```{r}
result$p_grid[which.max(result$posterior)]
```


```{r}
chainmode(samples, adj=0.01)
```

```{r}
mean(samples)
median(samples)
```

```{r}
sum(result$posterior * abs(result$p_grid - 0.5))
```

```{r}
loss = sapply(result$p_grid, function(d) sum(result$posterior * abs(d - p_grid)))

result$p_grid[which.min(loss)]
```

## p.63
### Dummy data
```{r}
dbinom(0:2, size=2, prob=6/9)
```

```{r}
dummy_w = rbinom(1e5, size=2, prob=6/9)
table(dummy_w)/1e5
```

```{r}
dummy_w = rbinom(1e5, size=9, prob=6/9)
simplehist(dummy_w, xlab="count")
```

## p.64
### Model checking

```{r}
result = globe_toss_uniform(6, 9, 1e5)
samples = sample(x=result$p_grid, prob=result$posterior, size=1e5, replace=T)

w = rbinom(1e5, size=9, prob=samples)
simplehist(w)
```


# Practice
## Easy
```{r}
p_grid = seq(from=0, to=1,length.out=1000)
prior = rep(1 , 1000)
likelihood = dbinom(6, size=9, prob=p_grid) 
posterior = likelihood * prior
posterior = posterior / sum(posterior)
set.seed(100)
samples = sample(p_grid, prob=posterior, size=1e4, replace=TRUE)
```
### 3E1
```{r}
mean(samples < 0.2)
```

### 3E2
```{r}
mean(samples < 0.8)
```

### 3E3
```{r}
mean(samples > 0.2 & samples < 0.8)
```

### 3E4
```{r}
quantile(samples, 0.2)
```

### 3E5
```{r}
quantile(samples, 0.8)
```

### 3E6
```{r}
HPDI(samples, 0.66)
```

### 3E6
```{r}
PI(samples, 0.66)
```

## Medium
```{r}
rm(.Random.seed, envir=globalenv())
```

### 3M1
```{r}
result = globe_toss_uniform(8, 15, 1e4)
plot(x=result$p_grid, y=result$posterior)
```

### 3M2
```{r}
samples = sample(x=result$p_grid, size=1e4, prob=result$posterior, replace=T)
HPDI(samples, 0.9)
```

### 3M3
```{r}
pred = rbinom(1e4, size=15, prob=samples)
mean(pred == 8)
```

### 3M4
```{r}
pred = rbinom(1e4, size=9, prob=samples)
mean(pred == 6)
```
```{r}
mean(samples > 0.65 & samples < 0.75)
```


### 3M5
```{r}
globe_toss_half_zero = function(x, N, num_points=1e4) {
  p_grid = seq(from=0, to=1, length.out=num_points)
  prior = rep(1, num_points)
  prior[p_grid < 0.5] = 0
  
  likelihood = dbinom(x, size=N, prob=p_grid)
  posterior = likelihood * prior
  posterior = posterior / sum(posterior)

  return(list(p_grid=p_grid, posterior=posterior))
}

result = globe_toss_half_zero(8, 15)
plot(x=result$p_grid, y=result$posterior)
```

```{r}
samples = sample(x=result$p_grid, size=1e4, prob=result$posterior, replace=T)
HPDI(samples, 0.9)

pred = rbinom(1e4, size=15, prob=samples)
mean(pred == 8)

pred = rbinom(1e4, size=9, prob=samples)
mean(pred == 6)
```

```{r}
mean(samples > 0.65 & samples < 0.75)
```

Using the second prior, we obtain higher probability around the true value of 0.7. That is, imbuing the model with prior knowledge enhanced our estimates.

### 3M6
```{r}
true_p = 0.7
num_toss = 2220

result = globe_toss_uniform(rbinom(1, num_toss, true_p), num_toss, 1e4)
samples = sample(x=result$p_grid, size=1e4, prob=result$posterior, replace=T)
pi = PI(samples, 0.99)
pi
pi[2] - pi[1]

# Test for different prior
result = globe_toss_uniform(rbinom(1, num_toss, true_p), num_toss, 1e4)
samples = sample(x=result$p_grid, size=1e4, prob=result$posterior, replace=T)
pi = PI(samples, 0.99)
pi
pi[2] - pi[1]
```

## Hard
```{r}
data(homeworkch3)
```

### 3H1
```{r}
num_boys = sum(birth1) + sum(birth2)
total = length(birth1) + length(birth2)
num_samples = 1e4

result = globe_toss_uniform(num_boys, total, num_samples)
plot(x=result$p_grid, y=result$posterior)

result$p_grid[which.max(result$posterior)]
```

### 3H2
```{r}
samples = sample(x=result$p_grid, size=num_samples, prob=result$posterior, replace=T)
HPDI(samples, 0.5)
HPDI(samples, 0.89)
HPDI(samples, 0.97)
```

### 3H3
```{r}
pred = rbinom(num_samples, size=total, prob=samples)
dens(pred)

abline(v=num_boys)
```

The model is quite precise in this case. Diverging just a little from the real value.

### 3H4
```{r}
num_boys1 = sum(birth1)
total = length(birth1)

pred = rbinom(num_samples, size=total, prob=samples)
dens(pred)
abline(v=num_boys1)
```

The model starts to diverge from the real data. That is, the model expected more boys on the first birth than we have in the dataset.

### 3H5
```{r}
num_girls_first = sum(birth1 == 0)
boys_followed_girls = birth2[birth1 == 0]

pred = rbinom(num_samples, size=num_girls_first, prob=samples)
dens(pred)
abline(v=sum(boys_followed_girls))
```


